### 2022-04-26 paper/Chapters/Background.tex

It doesn't make sense to build or purchase a powerful PC for use at home and another power laptop to handle working on the go, especially with rising prices, so many people are moving to exclusively powerful laptops or looking for other solutions.

>>>>

\section{Summary}\label{sec:BackgroundSummary}

Even after decades of technological innovation, every consumer computer is still a compromise between power, portability, and price.
The desktop computer has been the best option for high power machines with a lower overall cost, but lacks the portability that has become a necessity in modern times.
Laptops and gaming laptops have risen in popularity for their portability and flexibility, but their power comes at a cost premium that makes it inaccessible for many consumers.
While thin clients are a viable solution for the corporate world the power is provided by a dedicated server and end-user costs are kept at a minimum, their lack in portability highlights a gap in the currently available technology.
There currently isn't a solution for a cheap an portable device that can leverage the power of an existing computer.
Specifically, it doesn't make sense to build or purchase a powerful PC for use at home and another power laptop to handle working on the go.

### 2022-04-26 paper/Chapters/Evaluation.tex

\noindent
Once all the data is collected, the data from the host and client machines for a single trial are organized into a single CSV file for processing.
In order to ensure that the data is comparable, all extraneous factors must be removed from the data.
Namely, the network latency and difference in computer clocks must be removed.
By keeping the host and client computers attached to the same network through ethernet, the network latency is minimized.
But the task of keeping two computer's internal clocks synchronized is not trivial.
Even after instructing the operating system to synchronize its clock with an internet time server, the two computers may still report different times.
In some cases, the difference between the host and client computer's system time was noticed to be upwards of a full second after resynchronizing both clocks \cite{time.is}.
This is partially accounted for in step \emph{6} of the above process, and the remaining inaccuracy can now be removed by subtracting the time difference between the when the client presses the button and the time the host executes that command.
It is important to not subtract the entire round trip time of the input, as that is what is being testing in the following section, but rather just the time between when the client sends the input and when the host receives it.
This ensures that any inaccuracies in each computer's internal clock will not impact the data.

@@

\vspace{3mm} % Get the header of the block onto the next page. This can be removed if formatting changes
\begin{lstlisting}[style=plaintext,title=Statistics recorded while streaming KiCad (Figure \ref{fig:RealWorldKicad})]
Global Video Stats
----------------------------------------------------------
Incoming frame rate from network: 61.23 FPS
Decoding frame rate: 61.23 FPS
Rendering frame rate: 61.23 FPS
Frames dropped by your network connection: 0.00%
Average network latency: 1 ms (variance: 0 ms)
Average decoding time: 4.24 ms
Average rendering time (including monitor V-sync latency): 0.06 ms
Total dropped frames: 0
\end{lstlisting}

>>>>

\noindent
Once all the data is collected, the data from the host and client machines for a single trial are organized into a single CSV file for processing.
In order to ensure that the data is comparable, all extraneous factors must be removed from the data.
Namely, the network latency and difference in computer clocks must be removed.
By keeping the host and client computers attached to the same network through ethernet, the network latency is minimized.
But the task of keeping two computer's internal clocks synchronized is not trivial.

\begin{figure}[h]
  \startchronology[startyear=0,stopyear=10, height=0pt, dates=false, arrow=false, box=true]
  \chronograduation[event][dateselevation=0pt]{11}

  \chronoevent[date=false]{2}{Key is pressed on client}
  \chronoevent[date=false,markdepth=35pt]{6}{Key is received by host}
  \chronoevent[date=false]{8}{Key is pressed on host}

  \chronoperiode[textstyle=\raggedleft\colorbox{cyan!50}, color=cyan, startdate=false, bottomdepth=5pt, topheight=10pt, textdepth=-15pt,stopdate=false]{2}{6}{Network Latency}
  \chronoperiode[textstyle=\colorbox{green!50}, color=green, startdate=false, bottomdepth=5pt, topheight=10pt, textdepth=-15pt,stopdate=false]{6}{8}{Application Latency}
  \chronoperiode[textstyle=\colorbox{orange!50}, color=orange, startdate=false, bottomdepth=0pt, topheight=5pt, textdepth=-40pt,stopdate=false]{2}{8}{Total Latency}

  \stopchronology
  \caption[Timeline of a keystroke event]{Timeline of a keystroke event}
  \label{fig:KeystrokeTimeline}
\end{figure}

Even after instructing the operating system to synchronize its clock with an internet time server, the two computers may still report different times.
In some cases, the difference between the host and client computer's system time was noticed to be upwards of a full second after resynchronizing both clocks \cite{time.is}.
This is partially accounted for in step \emph{6} of the above process, but the remaining inaccuracy must be subtracted away.
Figure \ref{fig:KeystrokeTimeline} shows the timeline of a single keystroke event; after the client presses a key, it must first be sent and received by the host computer, and then the host must execute that input.
The time that this process takes is the total latency of the input.
For the purposes of evaluating the different applications, this delay must be split into the time it takes for the host to receive the input, called the network latency, and the time it takes for the host to actually execute the input, called the application latency.
Because the data loggers record the exact time that the test begins on both the client and host machines, the two times can be subtracted from each other to determine the network latency at the beginning of the test.
After recording the data, the total latency of each input recorded by the data loggers is subtracted by this value so that the time remaining is the application latency.
This ensures a fair and consistent comparison between each of the applications so that any inaccuracies in each computer's internal clock or the network speed has no impact on the data, and the only delay is that which is introduced by the application itself.

@@

\vspace{15mm} % Get the header of the block onto the next page. This can be removed if formatting changes
\begin{lstlisting}[style=plaintext,title=Statistics recorded while streaming KiCad (Figure \ref{fig:RealWorldKicad})]
Global Video Stats
----------------------------------------------------------
Incoming frame rate from network: 61.23 FPS
Decoding frame rate: 61.23 FPS
Rendering frame rate: 61.23 FPS
Frames dropped by your network connection: 0.00%
Average network latency: 1 ms (variance: 0 ms)
Average decoding time: 4.24 ms
Average rendering time (including monitor V-sync latency): 0.06 ms
Total dropped frames: 0
\end{lstlisting}

### 2022-04-26 paper/Chapters/Hardware.tex

When starting to choose parts for the hardware, there are many difficult questions to face.
First and foremost, the hardware had to be built with the time and resources available for this project.
Specifically, building a Single Board Computer from scratch would be completely infeasible, but the goal was also not to simply purchase an off the shelf SBC that was not tailored to the task at hand such as a traditional Raspberry Pi.
Instead, a better suited solution would be to utilize a \emph{Raspberry Pi Compute Module 4 (CM4)} which has \enquote{The power of Raspberry Pi 4 in a compact form factor for deeply embedded applications} \cite{rpi_cm4}.
As this perfectly lines up with the requirements for the project, this thesis will be built using the CM4.
However, due to it's small size and deeply embedded nature, the only way to communicate with the Compute Module is through two tiny mezzanine connectors as shown in Figure \ref{fig:rpi_cm4_mezzanine}.
With no other pin outs for anything such as power, HDMI, or USB, the rest of the hardware must be built on a custom Printed Circuit Board (PCB) that tailors the CM4 for this thesis.

@@

\begin{figure}[h]
  \centering
  %Height instead of width because the image is rotated
  \includegraphics[height=0.9\textwidth,angle=90]{Figures/smt_breakout}
  \captionsetup{width=.8\linewidth}
  \caption[SMD Breakout Board]{A breakout board similar to what would be needer for the USB controller used on the custom PCB}
  \label{fig:smt_breakout}
\end{figure}

@@

In order to produce a hardware device capable of fitting the needs of the project, a careful balance between power and cost had to be found.
The Raspberry Pi Compute Module 4 offered enough power to power the project while maintaining a low cost and slim profile that enabled it to fit in a small mobile device.
One the CM4 was chosen, a custom PCB had to be designed and manufactured to tailor the Compute module to the needs of the project.
This included choosing compatible electronic components, designing a board that provided the necessary functionality, and developing multiple iterations of the board to fix problems as they arose.

>>>>

When starting to choose parts for the hardware, there are many difficult questions to face.
First and foremost, the hardware had to be built with the time and resources available for this project.
Specifically, building a Single Board Computer from scratch would be completely infeasible, but the goal was also not to simply purchase an off the shelf SBC that was not tailored to the task at hand such as a traditional Raspberry Pi.
Instead, a better suited solution would be to utilize a \emph{Raspberry Pi Compute Module 4 (CM4)} which has \enquote{The power of Raspberry Pi 4 in a compact form factor for deeply embedded applications} \cite{rpi_cm4}.
Preliminary investigations showed that the CM4 is a promising candidate for the system and hence was chosen to be the starting point for the development of this project.
It's most important characteristic to keep in mind is due to it's small size and deeply embedded nature, the only way to communicate with the Compute Module is through two tiny mezzanine connectors as shown in Figure \ref{fig:rpi_cm4_mezzanine}.
With no other pin outs for anything such as power, HDMI, or USB, the rest of the hardware must be built on a custom Printed Circuit Board (PCB) that tailors the CM4 for this thesis.

@@

\begin{figure}[h]
  \centering
  %Height instead of width because the image is rotated
  \includegraphics[height=0.9\textwidth,angle=90]{Figures/smt_breakout}
  \captionsetup{width=.8\linewidth}
  \caption[SMD Breakout Board]{A breakout board similar to what would be needed for the USB controller used on the custom PCB}
  \label{fig:smt_breakout}
\end{figure}

@@

In order to produce a hardware device capable of fitting the needs of the project, a careful balance between power and cost had to be found.
The Raspberry Pi Compute Module 4 offered enough power to power the project while maintaining a low cost and slim profile that enabled it to fit in a small mobile device.
Once the CM4 was chosen, a custom PCB had to be designed and manufactured to tailor the Compute Module to the needs of the project.
This included choosing compatible electronic components, designing a board that provided the necessary functionality, and developing multiple iterations of the board to fix problems as they arose.
After three iterations, a board was finally produced that could be used to develop the software portion of the thesis.

### 2022-04-26 paper/Chapters/Software.tex

\todoadvisor{Summary}

>>>>

\section{Summary}\label{sec:SoftwareSummary}

In order to connect the client device to the host computer, a software solution had to be developed to compliment the hardware produced in the previous chapter.
Instead of reinventing the wheel and attempting to build a solution from scratch, focus was turned towards Nvidia's GameStream protocol, which has proven itself to be a reliable and capable solution that fits the needs of this thesis.
The open source project Moonlight has taken the closed source protocol and provides a usable API to build a custom client.
A fork of Moonlight was made for this thesis, and work was done to ease the development of the project on the \emph{ARM} architecture, as well as to build a suite of testing tools to evaluate the application against the existing solutions.
